Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2017-12-01 PAPER.tex
added normalization M's to equations

We are interested in quantifying how much variance in $\textbf{e}$ is lost after weighting $\textbf{r}$ and estimating the power 
spectrum according to OQE formalism. We investigate this by comparing two quantities we call the input power spectrum and 
output power spectrum: $P_{\rm in}$ and $P_{\rm out}$, defined as (ignoring normalization factors):

\begin{equation}
\label{eq:Pin}
P_{\rm in,\alpha} = \textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e}
\end{equation}

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{\rm out,\alpha} &=& \hat{\textbf{p}}_{r,\alpha}-\hat{\textbf{p}}_{x,\alpha} \nonumber \\
&=& \textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - \textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}

$P_{\rm in}$ represents the unweighted power spectrum of $\textbf{e}$, our simulated EoR signal, 
$\hat{\textbf{p}}_{x}$ is the power spectrum of the data alone, $\hat{\textbf{p}}_{r}$ is the power spectrum of the
data plus injection, and $P_{\rm out}$ is the difference between $\hat{\textbf{p}}_{r}$ and $\hat{\textbf{p}}_{x}$.

@@

One may wonder why $P_{\rm out}$ cannot be computed simply as the weighted power spectrum of $\textbf{e}$ alone, namely 
$P_{\rm out,\alpha} = \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e}$. In short, this naive 
expression fails to include correlations between the injected EoR signal and data, which emerges as statistical noise that is added to the 
estimate, decreasing the value of $P_{\rm out}$. Mathematically, we can show this by expanding out Equation \eqref{eq:sigloss}:

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &=& (\textbf{x}+\textbf{e})^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}(\textbf{x}+\textbf{e}) - 
\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x} \nonumber \\
&=& \textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} + \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} + \textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} - \textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}

@@

\begin{eqnarray}
\label{eq:pout_expand}
P_{\rm out, \alpha,\textbf{e} \gg \textbf{x}} &=& \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} + 
\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& \textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x}.
\end{eqnarray}

@@

\begin{eqnarray}
P_{\rm in,\alpha} &=& \textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e} \\
P_{\rm out,\alpha} &=& \textbf{r}^{\dagger}\textbf{C}_{r}^{-1}\textbf{Q}^{\alpha}\textbf{C}_{r}^{-1}\textbf{r} - \textbf{x}^{\dagger}
\textbf{C}_{x}^{-1}\textbf{Q}^{\alpha}\textbf{C}_{x}^{-1}\textbf{x} 
\end{eqnarray}

@@

Our revised, best to-date PAPER-64 power spectrum (using only one baseline separation type) is shown in Figure 
\ref{fig:ps1_data}. Again, black and gray points correspond to positive and negative power spectrum values respectively, with 
$2\sigma$ errors bars. Also plotted are the unweighted power spectrum upper limit (dashed blue) and theoretical prediction of 
noise (solid green). From this result, we quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, 
a higher limit than P64 by a factor of $\sim7$ in mK (though we only use one baseline type in our analysis as opposed to $3$). 
\cc{replace with kolopanis limit instead later}

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_reg.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_reg.png}
	\caption{Power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff}$. Black and gray points correspond to positive and 
negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted 
power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical $2\sigma$ noise level prediction based on 
observational parameters. This power spectrum result differs from P64 in that it only uses data from one type of baseline ($30$ m 
East/West baselines) instead of three. Major differences from previously published results stem from revisions regarding signal 
loss, bootstrapping, and the theoretical error computation. We quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at 
$k=0.34$ $h$Mpc$^{-1}$, a higher limit than P64 by a factor of $\sim7$ in mK \cc{replace with kolopanis limit later}.}
	\label{fig:ps1_data}
\end{figure*}

@@

As a consequence of our investigations, we have also revised power spectrum results from PAPER-64. For an analysis of only 
one baseline type, we quote a new $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than 
P64 by a factor of $\sim7$ in mK \cc{replace with kolopanis limit later}. The reasons for a previously underestimated limit and 
ways in which our new analysis differs can be summarized by the following:

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in P64, was underestimated by a factor of $\sim$$1000$ for inverse 
covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section 
\ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a 
factor of $\sim7$ mK higher than previous limits. Underestimated signal loss therefore represents the bulk of our revision. We 
note that PAPER's analysis is not the first to underestimate loss; results from the GMRT (\citealt{paciga_et_al2013}) were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were underestimated by a factor of $\sim$$2$ due to 
oversampling data whose effective number of independent samples was reduced from fringe-rate filtering. They were also 
overestimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power 
spectra (Section \ref{sec:Boot}). Hence, the net change for our error estimation is negligible, though the rationale behind the 
previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim$$2$ in mK (Section \ref{sec:PSSense}). We have verified our revised 
prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our 
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in P64 was too high for its data 
points.
\end{itemize}

>>>>

We are interested in quantifying how much variance in $\textbf{e}$ is lost after weighting $\textbf{r}$ and estimating the power 
spectrum according to OQE formalism. We investigate this by comparing two quantities we call the input power spectrum and 
output power spectrum: $P_{\rm in}$ and $P_{\rm out}$, defined as:

\begin{equation}
\label{eq:Pin}
P_{\rm in,\alpha} = \textbf{M}_{\rm in}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e}
\end{equation}

@@

\begin{eqnarray}
\label{eq:sigloss}
P_{\rm out,\alpha} &=& \hat{\textbf{p}}_{r,\alpha}-\hat{\textbf{p}}_{x,\alpha} \nonumber \\
&=& \textbf{M}_{rr}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r} - \textbf{M}_{xx}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where all $\textbf{M}$'s are normalization factors. $P_{\rm in}$ represents the unweighted power spectrum of $\textbf{e}$, our simulated EoR signal, 
$\hat{\textbf{p}}_{x}$ is the power spectrum of the data alone, $\hat{\textbf{p}}_{r}$ is the power spectrum of the
data plus injection, and $P_{\rm out}$ is the difference between $\hat{\textbf{p}}_{r}$ and $\hat{\textbf{p}}_{x}$.

@@

One may wonder why $P_{\rm out}$ cannot be computed simply as the weighted power spectrum of $\textbf{e}$ alone, namely 
$P_{\rm out,\alpha} = \textbf{M}_{ee}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e}$. In short, this naive 
expression fails to include correlations between the injected EoR signal and data, which emerges as statistical noise that is added to the 
estimate, decreasing the value of $P_{\rm out}$. Mathematically, we can show this by expanding out Equation \eqref{eq:sigloss}:

\begin{eqnarray}
\label{eq:crossterm}
P_{\rm out,\alpha} &=& \textbf{M}_{rr}(\textbf{x}+\textbf{e})^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}(\textbf{x}+\textbf{e}) - 
\textbf{M}_{xx}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x} \nonumber \\
&=& \textbf{M}_{rr}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} + \textbf{M}_{rr}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}
^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
&+& \textbf{M}_{rr}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} + \textbf{M}_{rr}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x} \nonumber \\
&-& \textbf{M}_{xx}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}
^{\alpha}\textbf{R}_{x}\textbf{x}.
\end{eqnarray}

@@

\begin{eqnarray}
\label{eq:pout_expand}
P_{\rm out, \alpha,\textbf{e} \gg \textbf{x}} &=&  \textbf{M}_{rr}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} +
 \textbf{M}_{rr}\textbf{x}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{e} \nonumber \\
 &+&  \textbf{M}_{rr}\textbf{e}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{x}.
\end{eqnarray}

@@

\begin{eqnarray}
P_{\rm in,\alpha} &=&  \textbf{M}_{\rm in}\textbf{e}^{\dagger}\textbf{I}\textbf{Q}^{\alpha}\textbf{I}\textbf{e} \\
P_{\rm out,\alpha} &=&  \textbf{M}_{rr}\textbf{r}^{\dagger}\textbf{C}_{r}^{-1}\textbf{Q}^{\alpha}\textbf{C}_{r}^{-1}\textbf{r} \nonumber \\
&&\phantom{=====} -  \textbf{M}_{xx}\textbf{x}^{\dagger}\textbf{C}_{x}^{-1}\textbf{Q}^{\alpha}\textbf{C}_{x}^{-1}\textbf{x} 
\end{eqnarray}

@@

Our revised, best to-date PAPER-64 power spectrum (using only one baseline separation type) is shown in Figure 
\ref{fig:ps1_data}. Again, black and gray points correspond to positive and negative power spectrum values respectively, with 
$2\sigma$ errors bars. Also plotted are the unweighted power spectrum upper limit (dashed blue) and theoretical prediction of 
noise (solid green). From this result, we quote a best $2\sigma$ upper limit of $(242.5$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, 
a higher limit than P64 by a factor of $\sim10$ in mK (though we only use one baseline type in our analysis as opposed to $3$). 
\cc{replace with kolopanis limit instead later}

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_reg.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_reg.png}
	\caption{Power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff}$. Black and gray points correspond to positive and 
negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted 
power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical $2\sigma$ noise level prediction based on 
observational parameters. This power spectrum result differs from P64 in that it only uses data from one type of baseline ($30$ m 
East/West baselines) instead of three. Major differences from previously published results stem from revisions regarding signal 
loss, bootstrapping, and the theoretical error computation. We quote a best $2\sigma$ upper limit of $(242.5$ mK$)^{2}$ at 
$k=0.34$ $h$Mpc$^{-1}$, a higher limit than P64 by a factor of $\sim10$ in mK \cc{replace with kolopanis limit later}.}
	\label{fig:ps1_data}
\end{figure*}

@@

As a consequence of our investigations, we have also revised power spectrum results from PAPER-64. For an analysis of only 
one baseline type, we quote a new $2\sigma$ upper limit of $(242.5$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than 
P64 by a factor of $\sim10$ in mK \cc{replace with kolopanis limit later}. The reasons for a previously underestimated limit and 
ways in which our new analysis differs can be summarized by the following:

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in P64, was underestimated by a factor of $\sim$$1000$ for inverse 
covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section 
\ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a 
factor of $\sim10$ mK higher than previous limits. Underestimated signal loss therefore represents the bulk of our revision. We 
note that PAPER's analysis is not the first to underestimate loss; results from the GMRT (\citealt{paciga_et_al2013}) were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were underestimated by a factor of $\sim$$2$ due to 
oversampling data whose effective number of independent samples was reduced from fringe-rate filtering. They were also 
overestimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power 
spectra (Section \ref{sec:Boot}). Hence, the net change for our error estimation is negligible, though the rationale behind the 
previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim$$2$ in mK (Section \ref{sec:PSSense}). We have verified our revised 
prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our 
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in P64 was too high for its data 
points.
\end{itemize}
